Deployed artificial-intelligence agents are increasingly governed by a stated
natural-language purpose, yet no existing tool measures whether an agent actually
pursues that purpose. Agent red-teaming probes systems against fixed attack
catalogues, and inverse reinforcement learning (IRL) reconstructs an agent's
implicit objective only at the model-weight or training level, which closed
deployments do not expose. We introduce the Inverse-Objective Regulator (IOR), a
black-box system that infers a revealed objective from an agent's tool-call
trajectories using Bayesian IRL, computes its divergence from the declared
purpose over a shared feature basis, and steers adversarial probes at the
dimensions of highest divergence. The inference-probe feedback loop is the core
contribution: probe results refine the inferred objective, which in turn targets
the next round of probes. A Goal Decomposition Judge bridges natural-language
purposes into the feature space IRL requires without access to model internals,
and a pluggable feature basis spans a deterministic model-free floor, a single
language-model judge, and a judge ensemble with quantified inter-judge agreement.
We describe the full audit protocol and contribute an open gym of three
planted-divergence reference environments for evaluating it. As an initial result,
a deterministic model-free feature basis predicts held-out agent behaviour with a
mean log-loss reduction of $0.649$ over a uniform baseline. A column-mean ablation
probes what the Bayesian IRL adds beyond an LLM judge: on a deterministic basis the
IRL divergence ranking is correlated with but not reducible to the mean (Spearman
$0.77$, top-dimension agreement $33\%$), the comparison is itself sensitive to the
simplex normalisation, and BIRL additionally supplies the posterior uncertainty and
full-mode counterfactual reasoning that a mean cannot. Divergence recovery with an
Opus~4.8 judge is strong where the divergence is encoded in the trace (Pearson
$r \approx 0.97$ on a cost-minimisation agent) but fails where it is not (a
faithfulness agent whose logged observations omit the content signal), which
delineates the method's scope. The targeted-probing and closed-loop evaluations
remain specified as protocols. Findings are designed to map onto OWASP,
MITRE ATLAS, and NIST AI RMF taxonomy nodes so that output is legible to regulators
auditing intended-purpose conformity under the EU AI Act.
